---
title: "Bayesian Mixture Model - Stan Model"
author: "Mehrsa Khaleghikarahrodi & Gretchen A. Macht"
date: '2023-05-25'
output: word_document
---
# (1) Stan is writing a program that reflects the model that one designs which relates inputs to outputs for producing estimates. 

# Install packages
```{r}
library(rstan)
```

# (2) Stan is taking inputs to estimate outputs. Refer to equation (3) in the paper.
# (2a - data) The name and format (integer or vector) of the inputs are specified. in addition to specifying the data, the priors for conducting the analysis need to be specified.
# (2b - parameters) Refers to the estimators (outputs) that the model generates.
# (2c - model) The model is the combination of inputs and how it establishes a relation with outputs to produce the estimates.
# The model is a "Multivariate Model" that accounts for the correlation between shape parameters (i.e., alpha & beta) & includes random effect to allow for individual influence on the fitting of distributions.

# Bayesian Multivariate Mixture Model + random effects 
```{r}
Model= 
"data {
int N; //number of observations
vector[N] y; //this will be either the shape1 & shape2 parameter
int M; //number of individuals
vector[N] x; //parameter identifier for alpha & beta
int g[N] ; //the individual identifier

real beta_a; //this will be the prior a shape parameter for the beta distribution that determines the proportion of each group
real beta_b; //this will be the prior b shape parameter for the beta distribution that determines the proportion of each group

real prior1_mean1; //this will be the prior mean for the first group
real prior1_mean2; //this will be the prior mean for the second group

real prior1_variance1; //this will be the prior variance of the mean for the first group
real prior1_variance2; //this will be the prior variance of the mean for the second group

real prior2_mean1; //this will be the prior mean for the first group
real prior2_mean2; //this will be the prior mean for the second group

real prior2_variance1; //this will be the prior variance of the mean for the first group
real prior2_variance2; //this will be the prior variance of the mean for the second group

real prior1_variance_a1; //this will be the prior variance shape parameter a for the first group
real prior1_variance_b1; //this will be the prior variance shape parameter b for the first group

real prior1_variance_a2; //this will be the prior variance shape parameter a for the second group
real prior1_variance_b2; //this will be the prior variance shape parameter b for the second group

real prior2_variance_a1; //this will be the prior variance shape parameter a for the first group
real prior2_variance_b1; //this will be the prior variance shape parameter b for the first group

real prior2_variance_a2; //this will be the prior variance shape parameter a for the second group
real prior2_variance_b2; //this will be the prior variance shape parameter b for the second group

real prior_variance_aID; //this will be the prior variance shape parameter a for the first group
real prior_variance_bID; //this will be the prior variance shape parameter b for the first group
}

//this related to posterior values
parameters {
  ordered[2] mu_alpha; //this will be the means for the distributions
  real sigma_alpha[2]; //this will be the variances for the distributions
  ordered[2] mu_beta; //this will be the means for the distributions
  real sigma_beta[2]; //this will be the variances for the distributions
  real<lower=0, upper=1> theta;
  real <lower=0> sigma_id;
  real a[M]; //the random effects
}

model {
sigma_alpha[1] ~ gamma(prior1_variance_a1, prior1_variance_b1);
sigma_beta[2] ~ gamma(prior1_variance_a2, prior1_variance_b2);

sigma_alpha[1] ~ gamma(prior2_variance_a1, prior2_variance_b1);
sigma_beta[2] ~ gamma(prior2_variance_a2, prior2_variance_b2);

mu_alpha[1] ~ normal(prior1_mean1, prior1_variance1);
mu_beta[2] ~ normal(prior1_mean2, prior1_variance2);

mu_alpha[1] ~ normal(prior2_mean1, prior2_variance1);
mu_beta[2] ~ normal(prior2_mean2, prior2_variance2);

sigma_id ~ gamma(prior_variance_aID, prior_variance_bID);
a ~ normal(0,sigma_id);

//the probability of each distribution. it is beta because we are looking at the probability
theta ~ beta(beta_a, beta_b);

//this line compiles all of the given the probability and normal distribution values
for (n in 1:N)
   target += log_mix(theta,
                     normal_lpdf(y[n] | mu_alpha[1]*x+mu_beta[1]*(1-x)+a[g[N]], sigma_alpha[1]*x+sigma_beta[1]*(1-x)),
                     normal_lpdf(y[n] | mu_alpha[2]*x+mu_beta[2]*(1-x)+a[g[N]], sigma_alpha[2]*x+sigma_beta[2]*(1-x)));
}
"
```

# (3) This section takes the designed model and the specifications mentioned above (given the data and prior information) to produce the estimates. There are two sets of prior information one being informative and the other being non-informative. 
# All users’ shape parameters of Beta distribution charging profiles are attained to initiate and execute the mixture modeling through the prior knowledge adapted from Khaleghikarahrodi and Macht (2023b). Given the summary of statistics on users with procrastination-like charging behavior, as outlined in Khaleghikarahrodi and Macht (2023b), the method of moments is utilized to infer the shape parameters for the prior information.

# Creating the data file for Multivatiate Model
# (3a - non-informative prior) The wider the selected variance, the more uninformative the prior (i.e., variance=100).
```{r non-informative prior [26%]} 
mod_data_beta=list(N=nrow(Multi.Beta), #N
              y=Multi.Beta$Combo.Shape, #y
              M=length(unique(Multi.Beta$Identifier)),
              x=Multi.Beta$Shape.ID, #x
              g=Multi.Beta$Identifier,
              
              beta_a=1,
              beta_b=1,
              
              prior1_mean1=3.576527, 
              prior1_variance1=1000, 
              prior1_mean2=1.787060,
              prior1_variance2=1000,
              
              prior2_mean1=1.719126, 
              prior2_variance1=1000, 
              prior2_mean2=2.509360,
              prior2_variance2=1000,
              
              prior1_variance_a1=1000,
              prior1_variance_b1=1000,
              prior1_variance_a2=1000,
              prior1_variance_b2=1000, 
              
              prior2_variance_a1=1000,
              prior2_variance_b1=1000,
              prior2_variance_a2=1000,
              prior2_variance_b2=1000) 

fit_shape=stan(model_code=Model,
         model_name="Model",
         data=mod_data_beta,
         iter=10000, chains=1, seed=100)

fit_shape

traceplot(fit_shape)
```

# (3b - informative prior) The informative prior includes more percise variance which is adpated from Khaleghikarahrodi and Macht (2023b). 
```{r informative prior} 
mod_data_beta.adj.v=list(N=nrow(Multi.Beta),#N
              y=Multi.Beta$Combo.Shape,#y
              M=length(unique(Multi.Beta$Identifier)),
              x=Multi.Beta$Shape.ID,#x
              g=Multi.Beta$Identifier,
              
              beta_a=1,
              beta_b=1,
              
              prior1_mean1=3.576527, 
              prior1_variance1=(6.706670^2), 
              prior1_mean2=1.787060,
              prior1_variance2=(1.175943^2),
              
              prior2_mean1=1.719126, 
              prior2_variance1=(3.231138^2), 
              prior2_mean2=2.509360,
              prior2_variance2=(1.787653^2),
              
              prior1_variance_a1=67.5,
              prior1_variance_b1=89.958832,
              prior1_variance_a2=191.5,
              prior1_variance_b2=2.765682, 
              
              prior2_variance_a1=67.5,
              prior2_variance_b1=20.880506,
              prior2_variance_a2=191.5,
              prior2_variance_b2=6.391408) 

fit_shape.adj.v=stan(model_code=Model,
         model_name="Model",
         data=mod_data_beta.adj.v,
         iter=10000, chains=1, seed=100)

fit_shape.adj.v

traceplot(fit_shape)
```
